---
title: "US-Apartment-Market-Insights"
output:
  pdf_document: 

  html_document: default
date: "2023-10-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
#install.packages("tidyverse")



```{r}

# Load necessary libraries
library(ggplot2)
library(readxl)
library(tidyverse)



```



```{r}

# Define file path based on current directory
file_path <- file.path(getwd(), "apartments_for_rent_classified_10K.csv")

# Check if file exists
if (!file.exists(file_path)) {
  stop("File not found in the current working directory!")
}

# Read the csv file
apartments_data <- read.csv(file_path, header = TRUE,sep = ';')

```

```{r}
# Display the first few rows of the dataset
head(apartments_data)

# Display the last few rows of the dataset
tail(apartments_data)

# Get the structure of the dataset
str(apartments_data)

# Get a summary of the dataset
summary(apartments_data)

# Understand the dimensions of the data
dim(apartments_data)

# Display the names of the columns
names(apartments_data)


```
```{r}
# Check for missing values in the dataset
sum(is.na(apartments_data))

# Remove rows with missing values (you can also impute them if you prefer)
apartments_data_clean <- na.omit(apartments_data)

# Check for duplicates
sum(duplicated(apartments_data_clean))

# Remove duplicates
apartments_data_clean <- unique(apartments_data_clean)

```


```{r}
head(apartments_data_clean)


# Histogram for numerical variables. For example, if you have a column named 'price':
ggplot(apartments_data_clean, aes(x = price)) + 
  geom_histogram(binwidth = 50, fill = "blue", alpha = 0.7) + 
  theme_minimal() + 
  labs(title = "Distribution of Price", x = "Price", y = "Frequency")

# Boxplot for numerical variables against categorical ones. E.g., 'price' against 'location (State)':
ggplot(apartments_data_clean, aes(x = state, y = price)) + 
  geom_boxplot() + 
  theme_minimal() + 
  labs(title = "Price Distribution by Location", x = "State", y = "Price")

# Scatterplot for two numerical variables. E.g., 'price' against 'square_feet':
ggplot(apartments_data_clean, aes(x = square_feet, y = price)) + 
  geom_point(alpha = 0.5) + 
  theme_minimal() + 
  labs(title = "Price vs. Square Feet", x = "Square Feet", y = "Price")


```


Handling Outliers
```{r}
# For a variable like 'price', compute the IQR:
Q1 <- quantile(apartments_data_clean$price, 0.25)
Q3 <- quantile(apartments_data_clean$price, 0.75)
IQR <- Q3 - Q1

# Define bounds for outliers
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Filter out outliers
apartments_data_no_outliers <- apartments_data_clean %>%
  filter(price >= lower_bound & price <= upper_bound)

# Visualize price without outliers
ggplot(apartments_data_no_outliers, aes(x = price)) + 
  geom_histogram(binwidth = 50, fill = "blue", alpha = 0.7) + 
  theme_minimal() + 
  labs(title = "Distribution of Price (No Outliers)", x = "Price", y = "Frequency")

```


1.  Are there specific amenities that significantly influence the classification category of an apartment listing?



a. Data Preparation: Extract and format amenities data.

b. Chi-Square Tests: For each amenity, conduct a Chi-Square test with the apartment category.

c. Visualization: Visualize the significant amenities in relation to apartment categories.

```{r}
# 1. Data Preparation
# Assuming amenities are separated by commas in the 'amenities' column
amenities_list <- strsplit(as.character(apartments_data$amenities), split = ",")
unique_amenities <- unique(unlist(amenities_list))

# Create a data frame where each amenity is a binary column
amenities_df <- matrix(0, nrow(apartments_data), length(unique_amenities))
colnames(amenities_df) <- unique_amenities

for (i in 1:length(amenities_list)) {
    amenities_df[i, amenities_list[[i]]] <- 1
}

amenities_df <- as.data.frame(amenities_df)

# 2. Chi-Square Tests
chi_square_results <- sapply(colnames(amenities_df), function(amenity) {
    chisq_test <- chisq.test(amenities_df[[amenity]], apartments_data$category)
    return(chisq_test$p.value)
})



# Combine results
results <- list(chi_square_results = chi_square_results)

results

df_chi_square <- data.frame(amenity = names(chi_square_results),
                            p_value = as.numeric(chi_square_results))

# 2. Plotting the p-values

threshold <- 0.05

ggplot(df_chi_square, aes(x = reorder(amenity, p_value), y = p_value)) + 
  geom_bar(stat="identity", aes(fill = p_value < threshold)) +
  coord_flip() + 
  geom_hline(yintercept = threshold, linetype="dashed", color = "red") +
  labs(title = "Significance Levels (p-values) of Amenities from Chi-Square Tests",
       x = "Amenity",
       y = "p-value",
       fill = "Significant at 0.05 level") +
  theme_minimal() +
  theme(legend.position="none")








```
In summary:

If an amenity's bar does not cross the red dashed line, it means that there's a statistically significant association between that amenity and the classification category of the apartment. In other words, the presence or absence of that amenity has a significant effect on how the apartment is categorized.
Conversely, if the bar crosses the red line, there's no significant association between that amenity and the apartment's classification category, at least at the 0.05 significance level.
```{r}
library(ggplot2)

# Top 10 Cities by Average Rental Price
top_cities <- head(city_avg_price, 10)

ggplot(top_cities, aes(x = reorder(cityname, price), y = price)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_hline(yintercept = overall_avg_price, color = "red", linetype = "dashed") +
  coord_flip() +  # Flip the coordinates for a horizontal bar chart
  labs(title = "Top 10 Cities by Average Rental Price",
       x = "City", y = "Average Rental Price") +
  theme_minimal()

# Top 10 States by Average Rental Price
top_states <- head(state_avg_price, 10)

ggplot(top_states, aes(x = reorder(state, price), y = price)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  geom_hline(yintercept = overall_avg_price, color = "red", linetype = "dashed") +
  coord_flip() +  # Flip the coordinates for a horizontal bar chart
  labs(title = "Top 10 States by Average Rental Price",
       x = "State", y = "Average Rental Price") +
  theme_minimal()

```
2. Listing Sources Analysis: How do rental prices differ based on the source of the listing (e.g., RentLingo, Zillow, etc.), and can we identify any specific trends associated with each source?

a. Descriptive Statistics and Measures of Variance
We'll compute the mean, median, variance, and standard deviation of rental prices for each listing source.

b. Normality Test
We'll employ the Shapiro-Wilk test to understand if the rental prices are normally distributed for each listing source.

c. ANOVA Analysis
If the data is normally distributed, we'll use ANOVA to determine if there are statistically significant differences in rental prices across listing sources.

d. Visualization
A bar chart will help visualize the average rental prices by listing source.


```{r}

# a. Descriptive Statistics and Measures of Variance
source_grouped <- aggregate(price ~ source, data = apartments_data, FUN = mean)
source_stats <- cbind(source_grouped, 
                      median = tapply(apartments_data$price, apartments_data$source, median),
                      variance = tapply(apartments_data$price, apartments_data$source, var),
                      sd = tapply(apartments_data$price, apartments_data$source, sd))

# b. Normality Test
# Filtering and sampling for Shapiro-Wilk test
shapiro_test_results <- lapply(split(apartments_data$price, apartments_data$source), function(x) {
  sample_size <- length(x)
  if(sample_size < 3) {
    return(NA)
  } else if(sample_size > 5000) {
    x <- sample(x, 5000)  # Taking a random sample of 5000 data points
  }
  shapiro.test(x)$p.value
})

# Convert the results to a data frame for better visualization
shapiro_df <- data.frame(source = names(shapiro_test_results), p_value = unlist(shapiro_test_results))

# c. ANOVA Analysis
anova_result <- aov(price ~ source, data = apartments_data)
anova_summary <- summary(anova_result)

# d. Visualization
ggplot_plot <- ggplot(source_stats, aes(x = reorder(source, -price), y = price)) +
  geom_bar(stat = "identity", fill = "dodgerblue") +
  coord_flip() +
  labs(title = "Average Rental Price by Source", x = "Listing Source", y = "Average Rental Price") +
  theme_minimal() +
  geom_errorbar(aes(ymin = price - sd, ymax = price + sd), width = 0.5)

# Output the computed values and the ggplot object
list(source_stats = source_stats, shapiro_test = shapiro_df, anova_summary = anova_summary, plot = ggplot_plot)







```

